---
title: "NYU"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Datathon Tutorial

The aim of this tutorial is to get you familiarized with BigQuery to query/filter/aggregate/export data with R. If you are familiar with Python, we also recommend that you check out the [Python version of this tutorial](bigquery_colab.ipynb).

## Prerequisites

You will need to have a valid Google account to be able to log in to Google Cloud Platform. If you do not have one, you can create one at https://accounts.google.com. If you will be also accessing restricted datasets, you may need to notify datathon organizers to register your account for data access.

## Setup

First, you need to run some initialization code. You can run the following cell by either pressing "Ctrl+Shift+Enter", or selecting "Cell -> Run cell and select below" menu, or clicking the equivalent button in the tool bar.

```{r}
# Install the "big R query" package, if neccessary by uncommenting the following two lines:
# install.packages('devtools')
# devtools::install_github("rstats-db/bigrquery")
library("bigrquery")
# Install ggplot2, uncomment next line if this is the first time this section is run.
# install.packages("ggplot2")
library("ggplot2")
# Re-install curl to avoid errors like:
# Error in curl::curl_fetch_memory(url, handle = handle) :
#   Error in the HTTP2 framing layer
# Uncomment next line if this is the first time this section is run.
# install.packages("curl")
# Install missing dependency, uncomment next line if this is the first time this section is run.
# install.packages("readr")
# Shared project. If you do not have the shared datathon project, you may use your own GCP project.
project_id <- "physionet-data"
options(httr_oauth_cache=FALSE)
# Wrapper for running BigQuery queries.
run_query <- function(query){
    data <- query_exec(query, project=project_id, use_legacy_sql = FALSE, max_pages = Inf)
    return(data)
}
```

## Analysis

```{r}
df <- run_query("
WITH patients AS (
  SELECT ad.SUBJECT_ID, ad.HADM_ID, ad.HOSPITAL_EXPIRE_FLAG, icd.ICD9_CODE FROM `mimiciii_clinical.admissions` AS ad
  INNER JOIN `mimiciii_clinical.diagnoses_icd` AS icd
  ON ad.SUBJECT_ID = icd.SUBJECT_ID
  WHERE icd.ICD9_CODE = '4275' OR icd.ICD9_CODE = 'V1253'),
arrested AS (
  SELECT p.SUBJECT_ID, p.HADM_ID, p.HOSPITAL_EXPIRE_FLAG, icd.ICD9_CODE FROM patients AS p
  LEFT JOIN `mimiciii_clinical.diagnoses_icd`  AS icd
  ON p.SUBJECT_ID = icd.SUBJECT_ID)
SELECT a.*, icd.SHORT_TITLE, icd.LONG_TITLE FROM arrested AS a
LEFT JOIN `mimiciii_clinical.d_icd_diagnoses` AS icd
ON a.ICD9_CODE = icd.ICD9_CODE")

```

```{r}
str(df)
```

```{r}
colnames(df)
```

```{r warnings=FALSE, message=FALSE}
library(psych)
describe.by(df)
```

The above information appears to be meaningles in regards to this particular type of analysis.

```{r}
paste0("Total number of patients with cardiac arrest ICD-9: ", nrow(df))
```

```{r}
library(tidyverse)
table(df$HOSPITAL_EXPIRE_FLAG)
```

* 20788 died while in the hospital.
* 145828 survived to out of hospital.

```{r}
library(ggplot2)
ggplot(df, aes(HOSPITAL_EXPIRE_FLAG)) + geom_bar(fill="lightblue") +
  xlab("0: Survived to Hospital Discharge, 1: Died in Hospital") + ylab("")
```

How many unique types of ICD9s in this dataset?

```{r}
length(unique(df$ICD9_CODE))
```

There are 2340 unique ICD codes.

How many different unique ICD codes exist for those who survived and those who did not?

```{r}
df %>% 
  group_by(HOSPITAL_EXPIRE_FLAG) %>% 
  summarise(ICD_Count = length(unique(ICD9_CODE)))
```

There is no obvious difference in count by the ICD_Count by those who survived and didn't survive.

Listing some of the ICD_Codes by whether the patient expired or not expired. Listing only 20 out of few thousand.

```{r}
a <- df %>% group_by(HOSPITAL_EXPIRE_FLAG) %>% 
  select(HOSPITAL_EXPIRE_FLAG, SHORT_TITLE)

print("Survived in the hospital after cardiac arrest: ")
a %>% filter(HOSPITAL_EXPIRE_FLAG == 0) %>% head(20)
```

```{r}
print("Died in the hospital after cardiac arrest: ")
a %>% filter(HOSPITAL_EXPIRE_FLAG == 1) %>% head(20)
```

The problem here is that all of these patients had a carest arrest. 

There are too many different diagnoses, let us focus on some specific ICDs.

--ICD 4019,4011,4010,40509,40519, HTN
--ICD 5723 Portal HTN
--ICD 4160 Pulmonary HTN
--ICD V4581,4111,41181,4142,4144,41401 Coronary
--ICD 24971,24961,24951,24981,24941,24991,24911,24960,24950,24990,24970,24920,24921,24910,24930,24931,24980,24940,24900,24901,25043, 25042 Diabetes
--ICD 2722,2724 HLD
--ICD 40391,40301 Renal disease
--ICD 4928, 5182, 5181, 7702 COPD

```{r}
icd_codes <- c("4019", "4011", "4010", "40509", "40519","5723","4160","V4581", "4111", "41181", "4142", "4144", "41401", "24971", "24961", '24951','24981','24941','24991','24911','24960','24950','24990','24970','24920','24921','24910','24930','24931','24980','24940','24900','24901','25043','25042','2722','2724','40391','40301','4928','5182','5181','7702')

b <- df %>% 
  filter(ICD9_CODE %in% icd_codes) %>% 
  group_by(HOSPITAL_EXPIRE_FLAG, SHORT_TITLE) %>% 
  select(HOSPITAL_EXPIRE_FLAG, SHORT_TITLE)
```

```{r}
print("Expired in the Hospital after Cardiac Arrest: ")
c <- as.data.frame(table(b$HOSPITAL_EXPIRE_FLAG, b$SHORT_TITLE))
c
```

How many are in each group with this filtered subset?

```{r}
b %>% group_by(HOSPITAL_EXPIRE_FLAG) %>% 
  summarise(TotalPts = length(HOSPITAL_EXPIRE_FLAG))
```

```{r}
c$mean <- ifelse(c$Var1 == 0, c$Freq/8248, c$Freq/1376)

library(reshape2)
ggplot(c, aes(x=Var2,y=mean,fill=factor(Var1))) +
  geom_bar(stat="identity", position="dodge") +
  scale_fill_discrete(name="Survival",
                      breaks=c(0,1),
                      labels=c("Survived", "Expired")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + xlab("ICD9") + ylab("Percent Frequency with ICD Present")
```

```{r}
project_id <- "nyu-datathon"

meds <- run_query("select  p.subject_id, p.hadm_id, rx.drug
from Team7.prescriptions rx INNER JOIN Team7population.population_adm p
on p.subject_id=rx.subject_id and p.hadm_id = rx.hadm_id where rx.drug in ('Albumin 25%',
'Dobutamine Hcl',
'Epinephrine-Sodium Chloride',
'Hydrocortisone Acetate 10%  Foam',
'Hydrocortisone Oint 0.5%',
'Methylpre',
'Phenylephrine  0.05% Nasal Spray',
'Phenylephrine  0.5% Nasal Spray',
'PrednisoLONE Acetate 0.12% Ophth. Susp.',
'Prednisolone Acetate',
'Propofol (Diprivan)',
'Propofol Diprivan',
'Sodium Bicarbonate 4.2% (2.4mEq)',
'albumin',
'sodium bicarbonate',
'Albumin',
'Albumin ',
'Dexamethasone Ophthalmic Soln 0.1%',
'Hydrocortisone Acetate Suppository',
'Hydrocortisone Na Succ',
'Hydrocortisone Oint 1%',
'NEO*IV*Phenobarbital',
'Neo-Synephrine',
'Phenylephrine HCl 1%',
'Albumin 5% (2.5gm)',
'Hydrocortisone Cream 2.5%',
'MEMAntine',
'Magnesium Sulfate (Latex Free)',
'sodium bica',
'Ketamine Study Drug (*IND*)',
'Prednisolone Acetate 1% Ophth. Susp.',
'Hydrocortisone (Rectal) 2.5% Cream',
'sodium bicar',
'sodium bicarb',
'Albumin, Human',
'Hydrocortisone Cream 0.5%',
'Dexamethasone Sod Phosphate',
'Methylphenidate HCl',
'Methylprednisolone',
'Phenobarbital Sodium',
'Epinephrine Inhalation',
'Hydrocortisone Acetate Ointment 1%',
'Dobutamine HCl',
'Epinephrine 1:1000',
'Hydrocortisone Cream 1%',
'Hydrocortisone Na Succinate',
'Methylene Blue 1%',
'Hydrocortisone',
'PENTObarbital Sodium',
'PrednisoLONE Acetate 1% Ophth. Susp.',
'Desmopressin Nasal',
'Albumin 5%',
'Magnesium Sulfate Replacement (Oncology)',
'Methylprednisolone Sodium Succ',
'Albumin 5% ',
'Norepinephrine Bitartrate',
'Epinephrine HCl',
'Alteplase',
'Midodrine HCl',
'Hydrocortisone Study Drug (*IND*)',
'Albumin 5% (12.5 g)',
'Albumin 5% (12.5gm)',
'Magnesium Citrate',
'Dobutamine',
'Albumin 5% (25 g)',
'Albumin 25% (12.5gm)',
'Albumin 25% (12.5 g)',
'EPINEPHrine',
'Dexmedetomidine HCl',
'Fludrocortisone Acetate',
'Methylprednisolone Na Succ',
'Midodrine',
'Dexmedetomidine',
'DOBUTamine',
'Calcium Chloride',
'Nitroglycerin SL',
'Nitroprusside Sodium',
'Thiamine',
'PHENYLEPHrine',
'Methylprednisolone Na Succ.',
'Albumin 5% (12.5g / 250mL)',
'Dexamethasone',
'MethylPREDNISolone Sodium Succ',
'Propofol (Generic)',
'Nesiritide',
'DopAmine',
'NORepinephrine',
'DOPamine',
'Albumin 5% (25g / 500mL)',
'Prednisone',
'PredniSONE',
'Magnesium Oxide',
'Albumin 25% (12.5g / 50mL)',
'Lidocaine',
'Epinephrine',
'Hydrocortisone Na Succ.',
'Milrinone',
'Phenylephrine HCl',
'Vasopressin',
'Phenylephrine',
'Nitroglycerin',
'Norepinephrine',
'Pantoprazole',
'Sodium Bicarbonate',
'Amiodarone HCl',
'Propofol',
'Amiodarone',
'Heparin',
'Calcium Gluconate',
'Magnesium Sulfate',
'SW')")
```

```{r}
labs <- run_query("WITH items as (select 
itemid, label, category 
from Team7population.d_items_subset
where itemid in (838    ,
226110    ,
225737    ,
225722    ,
8505    ,
227547    ,
227546    ,
227543    ,
8555    ,
6702    ,
6701    ,
220227    ,
223772    ,
3447    ,
226537    ,
223762    ,
225210    ,
227239    ,
220058    ,
220056    ,
225310    ,
225309    ,
225312    ,
645    ,
225698    ,
224828    ,
220235    ,
220224    ,
223830    ,
8502    ,
3312    ,
3313    ,
220621    ,
226107    ,
224291    ,
227292    ,
224289    ,
225664    ,
776    ,
777    ,
779    ,
778    ,
780    ,
1529    ,
224027    ,
679    ,
676    ,
811    ,
807    ,
223761    ,
224642    ,
677    ,
678    ,
220051    ,
220050    ,
220052    ,
456    ,
8441    ,
455    ,
52    ,
8368    ,
51    ))

SELECT 
ch.subject_id
, ch.hadm_id
, ch.icustay_id
, ch.itemid
, ch.charttime
, ch.value
, ch.valueuom
, ch.stopped
,i.label
,i.category
from Team7.chartevents ch join items i on i.itemid = ch.itemid")
```

Head of the Medication dataset.

```{r}
head(meds)
```

How many different drugs are there in this list?
Listed are 20 of the drugs.

```{r}
paste0("Number of unique drugs in the datsaet: ", length(unique(meds$drug)))
head(unique(meds$drug,20))
```

Merge the HOSPITAL_EXPIRED_FLAG to the medication dataset.

```{r}
colnames(meds) <- c("SUBJECT_ID", "HADM_ID", "DRUG")
df2 <- merge(df, meds, by=c("SUBJECT_ID","HADM_ID"))
head(df2)
```

Can we visualize a difference between steroid use in in-hospital survival?

```{r}
d <- df2 %>% select(HOSPITAL_EXPIRE_FLAG, DRUG)
head(d)
```

```{r}
steroid_list <- c("Dexamethasone", "Prednisone", "Methylprednisolone Na Succ.", "Methylprednisolone Na Succ", "Fludrocortisone Acetate", "Hydrocortisone Na Succinate", "Methylprednisolone Sodium Succ", "Prednisolone Acetate", "Methylprednisolone","Dexamethasone Sod Phosphate")

e <- d %>% filter(DRUG %in% steroid_list)
e
```

```{r}
dim(e)
```

26708 patients in this cohort.

```{r}
table(e$HOSPITAL_EXPIRE_FLAG)
ggplot(e, aes(HOSPITAL_EXPIRE_FLAG)) + geom_bar(fill="lightblue") +
  xlab("Whose who received steroids. 0: Survived in Hospital, 1: Expired in Hospital")
```

18284 patients died in the hospital compared to 8424 who survived.
Did the intensivists use steroids in these patients as a hail mary effort?

```{r}
library(kableExtra)
kable(as.data.frame(table(e$HOSPITAL_EXPIRE_FLAG, e$DRUG))) %>% 
  kable_styling(bootstrap_options = "striped", full_width = F)
```

Head of the Lab dataset:

```{r}
head(labs)
```

```{r}
unique(labs$label)
```

Found out that `labs` database is too large to work with at the moment.

Import in Demographics data in regards who died and survived to hospital discharge after cardiac arrest.

```{r}
df3 <- run_query("SELECT * FROM `Team7population.population_adm`")
```

```{r}
colnames(df3)
```

```{r}
library(corrplot)
df4 <- df3[,!colnames(df3) %in% c("subject_id", "hadm_id")]

library(fastDummies)
dum <- dummy_cols(df4[,c("marital_status", "ethnicity", "gender", "language", "religion")])

# Merging the dummy variables
df4 <- df4[,!colnames(df4) %in% c("marital_status", "ethnicity", "gender", "language", "religion")]
df4 <- cbind(df4, dum)
df4 <- df4[,!colnames(df4) %in% c("marital_status", "ethnicity", "gender", "language", "religion")]
head(df4)
```

```{r}
marital_status <- df4[,colnames(df4) %in% c("died", "age","marital_status_MARRIED" , "marital_status_SINGLE", "marital_status_WIDOWED", "marital_status_DIVORCED", "marital_status_SEPARATE", "marital_status_LIFE PARTNER")]
colnames(marital_status) <- c("Age", "Died", "Married", "Single", "Widowed", "Divorced", "Life Partner")
M <- cor(marital_status)
corrplot(M, order = "hclust")
```

```{r}
# Gender
gender <- df4[,colnames(df4) %in% c("died", "gender_M","gender_F")]
colnames(gender) <- c("Died", "Male", "Female")
M <- cor(gender)
corrplot(M, order = "hclust")
```

```{r}
# Religion
religion <- df4[,colnames(df4) %in% c("died", "religion_HINDU","religion_OTHER",    "religion_HEBREW","religion_JEWISH",                                    
"religion_MUSLIM",                                    
"religion_BAPTIST",                                   
"religion_BUDDHIST",                                  
"religion_CATHOLIC",                                  
"religion_EPISCOPALIAN",                              
"religion_GREEK ORTHODOX",                            
"religion_7TH DAY ADVENTIST",                         
"religion_JEHOVAH'S WITNESS",                         
"religion_PROTESTANT QUAKER",                         
"religion_CHRISTIAN SCIENTIST",                       
"religion_ROMANIAN EAST. ORTH",                       
"religion_UNITARIAN-UNIVERSALIST"    )]
colnames(religion) <- c("Died", "Hindu", "Other", "Hebrew", "Jewish", "Muslim", "Baptist", "Buddhist", "Catholic", "Episcopalian", "Greek Orthodox", "7th Day Adventist", "Jehovah's Witness", "Protestant Quaker", "Christian Scientist", "Romanian East. Orth", "Unitarian")
M <- cor(religion)
corrplot(M, order = "hclust")
```

```{r}
# Language
languages <- df4[,colnames(df4) %in% c("died"                                        
,"language_ENGL"                                      
,"language_HIND"                                      
,"language_SPAN"                                      
,"language_PTUN"                                      
,"language_ITAL"                                      
,"language_RUSS"                                      
,"language_CANT"                                      
,"language_PORT"                                      
,"language_VIET"                                      
,"language_PERS"                                      
,"language_AMER"                                      
,"language_SOMA"                                      
,"language_*BUR"                                      
,"language_KORE"                                      
,"language_CAPE"                                      
,"language_HAIT"                                      
,"language_TAGA"                                      
,"language_LAOT"                                      
,"language_BENG"                                      
,"language_GREE"                                      
,"language_FREN"                                      
,"language_MAND"                                      
,"language_*FUL"                                      
,"language_*ARM" )]

languages$foreign_language <- ifelse(languages$language_ENGL == 0, 1, 0)
languages <- languages[,colnames(languages) %in% c("died", "language_ENGL")]

colnames(languages) <- c("Died", "English")
M <- cor(languages)
M
corrplot(M, order = "hclust")
```

```{r}
# Died on the Y axis and Languages on the X axis
table(languages$Died, languages$English)
# Chi-sq test
chisq.test(languages$Died, languages$English, correct=FALSE)
```

```{r}
table(languages$Died)
```

```{r}
# Race
race <- df4[,colnames(df4) %in% c("died", 
                                    "ethnicity_WHITE"                                    
, "ethnicity_BLACK/AFRICAN AMERICAN"                   
, "ethnicity_ASIAN"                                    
, "ethnicity_ASIAN - ASIAN INDIAN"                     
, "ethnicity_HISPANIC OR LATINO"                       
, "ethnicity_PORTUGUESE"                               
, "ethnicity_WHITE - RUSSIAN"                          
, "ethnicity_ASIAN - VIETNAMESE"                       
, "ethnicity_ASIAN - CHINESE"                          
, "ethnicity_MULTI RACE ETHNICITY"                     
, "ethnicity_WHITE - BRAZILIAN"                        
, "ethnicity_HISPANIC/LATINO - PUERTO RICAN"           
, "ethnicity_BLACK/CAPE VERDEAN"                       
, "ethnicity_HISPANIC/LATINO - GUATEMALAN"             
, "ethnicity_HISPANIC/LATINO - DOMINICAN"              
, "ethnicity_BLACK/AFRICAN"                            
, "ethnicity_ASIAN - FILIPINO"                         
, "ethnicity_BLACK/HAITIAN"                            
, "ethnicity_NATIVE HAWAIIAN OR OTHER PACIFIC ISLANDER"
, "ethnicity_WHITE - OTHER EUROPEAN"                   
, "ethnicity_MIDDLE EASTERN"                           
, "ethnicity_ASIAN - OTHER"                            
, "ethnicity_ASIAN - JAPANESE"                         
, "ethnicity_WHITE - EASTERN EUROPEAN"                 
, "ethnicity_ASIAN - KOREAN"                           
, "ethnicity_AMERICAN INDIAN/ALASKA NATIVE"   )]

colnames(race) <- c("Died", "White", "Black", "Asian1", "Asian2", "Latino", "White2","White3", "Asian3", "Asian4", "Multi-Race", "Latino2", "Latino3", "Black2","Latino4", "Latino5", "Black3", "Asian4", "Black4","Asian5","White4","MiddleEastern","Asian5","Asian6","White5","Asian7","NativeAmerican")

race$White <- ifelse(race$White == 1 | race$White2 == 1 | race$White3 == 1 | race$White4 == 1 | race$White5 == 1, 1, 0)
race$Black <- ifelse(race$Black == 1 | race$Black2 == 1 | race$Black3 == 1 | race$Black4 == 1, 1, 0)
race$Asian <- ifelse(race$Asian1 == 1 | race$Asian2 == 1 | race$Asian3 == 1 | race$Asian4 == 1 | race$Asian5 == 1 | race$Asian6 == 1 | race$Asian7, 1, 0)
race$Latino <- ifelse(race$Latino == 1 | race$Latino2 == 1 | race$Latino3 == 1 | race$Latino4 == 1 | race$Latino5 == 1, 1, 0)

# race$`Multi-Race`
# race$MiddleEastern
# race$NativeAmerican
race <- race[,colnames(race) %in% c("Died", "White", "Black", "Asian", "Latino", "Multi-Race", "MiddleEastern", "NativeAmerican")]

M <- cor(race)
corrplot(M, order = "hclust")
M
```

```{r}
# Died on the Y axis and White Race on the X axis
table(race$Died, race$White)
# Chi-sq test
chisq.test(race$Died, race$White, correct=FALSE)
```

```{r}
# Died on the Y axis and Black Race on the X axis
table(race$Died, race$Black)

# Plot
z <- as.data.frame(table(race$Died, race$Black))
colnames(z) <- c("Died","Race","Freq")

ggplot(z, aes(x=Race,y=Freq,fill=factor(Died))) +
  geom_bar(stat="identity", position="dodge") +
  scale_fill_discrete(name="Survival",
                      breaks=c(0,1),
                      labels=c("Survived", "Expired")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + xlab("Black") + ylab("")

# Chi-sq test
chisq.test(race$Died, race$Black, correct=FALSE)
```

```{r}
# Died on the Y axis and Latino Race on the X axis
table(race$Died, race$Latino)

# Plotting
z <- as.data.frame(table(race$Died, race$Black))
colnames(z) <- c("Died","Race","Freq")

ggplot(z, aes(x=Race,y=Freq,fill=factor(Died))) +
  geom_bar(stat="identity", position="dodge") +
  scale_fill_discrete(name="Survival",
                      breaks=c(0,1),
                      labels=c("Survived", "Expired")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + xlab("Black") + ylab("")

# Chi-sq test
chisq.test(race$Died, race$Latino, correct=FALSE)
```

```{r}
# Died on the Y axis and Asian Race on the X axis
table(race$Died, race$Asian)
# Chi-sq test
chisq.test(race$Died, race$Asian, correct=FALSE)
```

```{r}
# Died on the Y axis and MiddleEastern Race on the X axis
table(race$Died, race$MiddleEastern)
# Chi-sq test
chisq.test(race$Died, race$MiddleEastern, correct=FALSE)
```

```{r}
# Died on the Y axis and NativeAmerican Race on the X axis
table(race$Died, race$White)
# Chi-sq test
chisq.test(race$Died, race$NativeAmerican, correct=FALSE)
```

```{r}
df_1 <- race
RACE <- colnames(df_1)[2:7][max.col(!is.na(df_1[,2:7]))]
races <- data.frame(DIED = df_1$Died, RACE = RACE)
head(races)
```

```{r}
kable(summary(races)) %>% 
  kable_styling(bootstrap_options = "striped", full_width = F)
races %>% group_by(RACE)# %>% summarise(Total_Died = sum(DIED)) %>% kable() %>% 
  #kable_styling(bootstrap_options = "striped", full_width = F)
```


